Draft Article

### 2019-12-20 article.tex
Finish most of aerial espionage

\begin{table}[ht]
\centering
\begin{tabular}{llr}
\textbf{Date}     & \textbf{\makecell[l]{U.S. Service \&\\ Aircraft Type}}   & \textbf{General Location} \\
8 April 1950      & USN PB4Y2 Privateer           & Baltic Sea                           \\
6 November 1951   & USN P2V Neptune               & Sea of Japan                         \\
13 June 1952      & USAF RB-29                    & Sea of Japan                         \\
7 October 1952    & USAF RB-29                    & East of Hokkaido/Kuril Is.           \\
29 July 1953      & USAF RB-50                    & Sea of Japan                         \\
4 September 1954  & USN P2V Neptune               & Sea of Japan                         \\
7 November 1954   & USAD RB-29                    & East of Hokkaido.Kuril Is.           \\
18 April 1955     & USAF RB-47                    & Off Kamchatka Peninsula              \\
10 September 1956 & USAF RB-50                    & Sea of Japan                         \\
2 September 1958  & USADF C-130                   & Soviet Armenia (near Turkish border) \\
1 May 1960        & CIA U-2                       & Sverdolsk, USSR                      \\
1 July 1960       & USAF RB-47                    & Barents Sea                          \\
10 March 1964     & USAF RB-66                    & East Germany
\end{tabular}
\caption{Summary of Soviet Shootdowns, 1950-1964}
\label{soviet-shootdowns}
\end{table}

@@

The general attitudes displayed in response to the Baltic Incident became the normal course of response to Soviet flights. The shootdown of the Navy Privateer was not an isolated incident; from then until the U-2 Incident in 1960, the United States lost nine aircraft to Soviet air defenses. Each time, the pattern of responses was the same---each side released dueling statements of protest, and the situation continued more or less as before. A B-29 that went missing on October 7, 1952 hit the front page of the \emph{New York Times} two days later. This mission is on Peterson's list as a ferret flight. Farquhar notes that the incident happened during the US election season and that the media thought the attack was intended to lower American prestige, but he does not mention any diplomatic consequences.\footcite[p.~43-44]{farquhar_aerial_2015} The official US response, however, is preserved on the State Department website. As with the Baltic Incident, the US simply demanded that the Soviets pay for the plane and return any survivors.\footcite{the_new_york_times_u.s._1952} Two more shootdowns in March 1953---one involving an American F–84 Thunderjet and the other an RAF Lincoln bomber---actually resulted in conciliatory responses from both sides, including a secret meeting to reduce aerial tensions \enquote{of which little is known.}\footcite[p.~45]{farquhar_aerial_2015}

@@

I contend that the lack of espionage equipment is precisely the reason the United States made a sustained public effort to recover the four airmen, which is something that almost never happened with ferret flights.\footnote{While some of the flights were shot down overseas, making recovery difficult, I am referring here to situations where the US was aware that the Soviets likely held their airmen in captivity and still made little effort to get them back.} Farquhar summarizes the American response to this incident as follows: ``Responding to the press attention, the Truman Administration acted swiftly, attempting to gain the fliers release through diplomatic pressure. The President ordered the Hungarian consulates in New York and Cleveland closed and banned private travel to the country. Legislatively, Truman asked Congress to pass a \$100 million Mutual Security Act to aid `selected persons residing in Soviet bloc states or refugees who wanted to form armed units' in opposition to Communism.''\footcite[p.~43]{farquhar_aerial_2015} Truman immediately took action against material Hungarian interests in the United States and threatened to empower \emph{armed dissidents} if the prisoners were not released. The contrast between this response and the one from just a year prior, when the US demanded \enquote{a prompt and thorough investigation} of the Baltic Incident, is stark.

@@

% \section{Civilian spy flights}
% \subsection{Project Aquatone}
% While Eisenhower was able to defuse some of the tension surrounding the military reconnaissance flights, even the most aggressive overflights could only hope to skirt the borders of Soviet territory; the United States couldn't just fly a B-50 over Moscow. What US intelligence-gathering capabilities needed was some way of photographing suspicious installations deep within Russia while still being able to claim the same benefit-of-the-doubt status that prevented even the most contentious overflights from escalating into a larger conflict. What they needed was a spy plane.

% The key feature of a spy plane is that it is designed to be minimally provocative. Any plane can be retrofitted with cameras or radios that allow it to perform useful reconnaissance functions---what separates a spy plane is that there can can be no doubt (in the mind of the target) that it is being used for anything \emph{but} reconnaissance. For the spy plane to serve its intended purpose as a minimally provocative agent of espionage then, two things must be true: its operation must be understood by adversaries to be non-military, and that distinction must meaningfully alter their threat perception and associated response. To build such a aircraft, Eisenhower approved Project Aquatone, the code name for the creation of the U-2.

% From its conception, the U-2 was purpose-built to be read as a civilian aircraft. Eisenhower required that the pilots of U-2 aircraft be CIA employees, specifically forbidding uniformed Air Force officers.\footcite[p.~33, Though many of the pilots did have Air Force backgrounds.]{lindgren_trust_2000} The CIA also had operational control of the program instead of the Air Force. \enquote{I want this whole thing to be a civilian operation,} Eisenhower wrote, when settling an operational dispute between the two departments. ``If uniformed personnel of the armed services of the United States fly over Russia, it is an act of war---legally---and I don't want any part of it.''\footcite[p.~60. The original source for this quote is an \emph{OSA History} that requires codeword clearance. It is quoted here by the History Staff of the CIA.]{pedlow_central_1992} This allowed the United States to \enquote{truthfully deny,} in the phrasing of the CIA, that any US \enquote{military planes} had flown over the USSR---which they had to do when the inevitable Soviet protest notes were filed after U-2 overflights began in 1956.\footcite[p.~109]{pedlow_central_1992}

% Having the CIA run the program---ostensibly for the purpose of deception---changed the nature of the program itself. If the only goal had been to achieve a level of plausible deniability in the event of a shootdown, then simply not wearing an Air Force uniform during the mission likely would have been sufficient. The \enquote{weather reconnaissance} excuse had even been used before with the ferret flights.\footcite[p.~45]{farquhar_aerial_2015} At the highest levels of American government, however, policymakers didn't just want to pretend that Aquatone was a civilian operation, they wanted it to \emph{be} a civilian operation---even if the NASA weather-craft cover didn't hold. \enquote{It is of utmost importance to differentiate in our minds, and to cause the Russians to differentiate in theirs, between Aquatone-type operations and reconnaissance by military aircraft} read a top-secret CIA memo from 1956.\footcite[p.~1]{miller_suggestions_1956} Both halves of that equation were equally important. The Russians must believe that these operations are peacetime civilian operations, and they must be correct about it.

% The United States was aware how provocative these reconnaissance flights might be, and Eisenhower was always concerned, correctly, about \enquote{the terrible propaganda impact that would be occasioned if a reconnaissance plane were to fail.}\footcite[p.~162]{pedlow_central_1992} Starting in 1958, the American administration ordered a drastic decrease in the number of overflights.\footcite[p.~51]{powers_operation_2004} But Eisenhower was also under tremendous political pressure to learn more about the Soviet missile program, as hawkish senators amplified fears of a so-called \enquote{missile gap,} concerned that Soviet ballistic missile capabilities were advancing faster than their own.\footcite[Fears of this missile gap quickly followed earlier fears of a \enquote{bomber gap,} which ironically the U-2 had been critical in disproving.]{licklider_missile_1970} He especially did not want to affect the upcoming Four Powers diplomatic summit in Paris, scheduled for the summer of 1960. Nonetheless, Eisenhower ordered two more overflights in 1960, one on April \nth{9} and the other on May \nth{1}. Only one of them returned.

% % It turned out however, that the Soviets opted to lodge their diplomatic protests privately, and soon they stopped protesting them at all.\footcite[p.~42]{lindgren_trust_2000}

% \subsection{The U-2 Incident}
% On the morning of May 1, 1960, Soviet Air Defense forces detected a high-altitude aircraft flying over Soviet Tajikistan. Though they were were not yet able to conclusively identify it, the foreign aircraft was an American U-2 spy plane performing an reconnaissance overflight of the USSR---at a time when the Soviets were deeply embarrassed by their inability to prevent such flights.\footcite{orlov_u-2_2007} The previous U-2 overflight on April 9, less than a month prior, had lasted a full six hours and the aftermath caused an upheaval in the Soviet military. Leadership ordered an investigative commission to root out the shortcomings of the Air Force and Air Defense agencies, and a series of charts were drawn up anticipating the routes of future U-2 flights. The next time one came through, the Soviets were absolutely determined to shoot it down.

% % \begin{figure}
% % \centering
% % \includegraphics[scale=0.3]{powers-flight-path.jpg}
% % \caption{Operation Grand Slam Flight Path}
% % \label{powers-flight-path}
% % \end{figure}

% The Soviets still proved unable to immediately stop the flight on May \nth{1}. A missile battalion in the plane's path was not on alert duty. Armed fighter aircraft were in the wrong positions. A frantic attempt to have another fighter plane literally ram it out of the sky was scrapped when the pilot failed to make visual contact.\footcite{orlov_u-2_2007} The highest possible levels of Soviet leadership were actively involved with the mission as it was taking place. A Soviet Colonel in the USSR Air Defense later recalled that Khrushchev ``clearly viewed the violation of their nation's skies by a foreign reconnaissance aircraft on the day of a Soviet national holiday, and just two weeks before a summit conference in Paris, as a political provocation.'' \footcite{orlov_u-2_2007}

% Why did the generally cautious Eisenhower authorize two overflights in such close proximity to a crucial summit? For one, the secrecy of the program worked against him. Eisenhower was not willing to make the U-2 program public, so he couldn't convince his critics that he had extremely high quality intelligence that no such missile gap existed, because to reveal the existence of his aerial photography would prompt questions about its origins.\footcite[p.~51]{lindgren_trust_2000} And the missions had been such a tremendous success that his reluctance to authorize more overflights grated his staff, especially the influential Director of Central Intelligence (DCI), Allen Dulles.\footcite[p.~354]{brugioni_eyes_2010}

% Almost four hours into the spy plane's flight on May 1, pilot Francis Gary Powers felt a dull \enquote{thump} and \enquote{a tremendous orange flash lit the cockpit and sky} as a Soviet surface-to-air (SAM) missile exploded behind him, ripping the wings off his plane and sending it into a tailspin.\footcite[p.~61]{powers_operation_2004} His subsequent capture by the Soviet military became an immediate international incident. Three days after NASA quietly reported that they had lost a U-2 type weather reconnaissance plane, Khrushchev announced that the USSR had shot down an American plane that had flown into their airspace, which he called \enquote{an aggressive provocation aimed at wrecking the Summit Conference.}\footcite[p.~112]{powers_operation_2004} Once Khrushchev gleefully debunked the resulting American cover story by revealing that they had taken Powers alive, Eisenhower admitted that the Gary Powers had been flying a spy plane, in a press conference where Eisenhower memorably declared that espionage was a \enquote{a distasteful but vital necessity} to prevent another attack like Pearl Harbor.\footcite{eisenhower_news_1960}

% % The contrast between the severity of response Khrushchev wanted to portray publicly and what the USSR actually did is highlighted by his own son Sergei, who edited and annotated the 2007 edition with notes featuring additional context for his father's statements.

% \section{Espionage in the open skies}
% \subsection{Minimal consequences with maxmimal noise}
% The U-2 Incident is often cited as a major failure of the Eisenhower administration. In the traditional telling, it significantly worsened US-Soviet relations at time when it seemed like lasting peace might be possible. I disagree; instead, I argue that that the U-2 Incident should be understood as an embarrassing but otherwise typical failure of an intelligence operation, one consistent with the minimal consequences typically applied to espionage.

% No one disputes that the collapse of the Paris Summit was directly attributable to the Gary Powers shootdown but, frankly, summits just aren't that important. From 1953 to the end of the Cold War in 1991, there were 23 total US-USSR summits.\footcite{fain_chronology_2011} The main issue that did not get resolved in 1960 was the status of Berlin. It also did not get resolved at the next summit in 1961, a comparatively warmer affair between Khrushchev and Eisenhower's successor, John F. Kennedy. If the worst that happened because of the U-2 incident was an unproductive summit meeting, then that is a relatively minor punishment. With the late rollout of the captured pilot, the dramatic scene at the summit, and the cancellation of Eisenhower's visit, Khrushchev publicly embarrassed the United States without meaningfully altering Soviet policy towards it.

% What would a major punishment have looked like? One proportional response Khrushchev could have considered was to force the issue on regional air bases. Since the US was using nearby bases to launch invasive reconnaissance missions, the USSR could have demanded that the US either close some of them or agree to halt surveillance flights. This likely would have involved some sort of trade, but it is not an unreasonable hypothetical. The Soviets were incredibly preoccupied with removing the forward-deployed Jupiter missiles from Turkey. Although the missiles were deployed to bases well within NATO territory, and arguably less threatening to Soviet security than American overflights of the Soviet bloc, the existence of these missile became a central bargaining chip in the Cuban Missile Crisis. By contrast, the Soviets simply did not have the same level of interest in curtailing reconnaissance flights. That both sides of the Cold War had strong incentives to avoid military escalation is well-documented, but at least threatening military retaliation for launching provocative military operations would not seem to have been out of place.

% While I argue that the response to the U-2 flight was within the reasonable bounds for espionage, I acknowledge that it was still unprecedented in the context of reconnaissance flights. There are a number of reasonable explanations for why the U-2 shootdown escalated. Partially the Soviets felt humiliated by a technology for which they had no countermeasures and the uniquely insulting way the overflights flew deep into Soviet territory. Eisenhower thought that Khrushchev might have been looking for an excuse to cancel his visit to Russia.\footcite[p.~555]{eisenhower_waging_1965} Others in his administration believed that the U-2 Incident was an excuse to justify an intensification of hostilities that had already been decided by hard-liners in the Kremlin.\footcite[p.~328]{kistiakowsky_scientist_1976}

% Most scholars who study this period argue that the cumulative tension created by these flights contributed to an overall chilly tone in US-Soviet relations. Farquhar, for instance, refers to this as the \enquote{cycle of hostility,} in which the series of aerial incidents increased suspicions of military buildup on both sides, leading to even more aerial incidents as both powers attempted to verify their suspicions.\footcite[p.~43]{farquhar_aerial_2015} While it's true that the there were many aerial incidents, it is not at all clear how this affected any other aspect of US-Soviet diplomacy. I can't definitively say that there wasn't some diplomatic issue that would have gotten resolved at a summit absent these reconnaissance flights, but I can't locate an obvious one either.

% Consider the period of time over which these flights took place. If aerial reconnaissance had risen to anything more than a lingering irritant, then it should have to have been resolved in some way before 1960. The US-Soviet relationship saw huge shifts, including periods of optimism, over the 11 years preceding the U-2 incident. During this period, hundreds of flights took place in contentious territory---many of them purposefully violating Soviet boundaries, causing some of them to be shot down---and the most significant consequence was a failed summit meeting, after which the flights continued on almost as before. Both sides were so willing to contain the issue that many of the airmen never saw their families again.

% \subsection{The U-2 after Powers}
% When the US really needed intelligence that only an overflight could provide, the diplomatic consequences that the USSR had imposed were not sufficient to discourage it. Two months to the day after Powers was shot down, a Soviet MiG-19 opened fire on an American RB-47H and captured the plane's navigator and co-pilot. While Powers was serving a 10-year sentence in Soviet prison, the two airmen were returned to the US in less than six months with only mild fanfare. Khrushchev wrote in his memoir that he wanted to continue \enquote{our general line of peaceful coexistence} and cited the resolution of this incident as an example of such, because the US was forced to make a formal request for the return of the airmen.\footcite[p.~256-257]{khrushchev_memoirs_2007} The only concessions the US made in return were to announce the discontinuation of its U-2 overflights (which Kennedy was already committed to) and to not make an issue of the illegal detention of the pilots.\footcite{time_cold_1961} For spy flights after Powers, the USSR simply took the necessary immediate countermeasures and shot down the planes that it could. Just as with traditional espionage, the countermeasures did not punish the attempt.

% Until the day he died, Khrushchev greatly exaggerated the lasting effect of the U-2 Incident. He wrote in his memoir that \enquote{the commander of American forces in West Germany gave the order not to fly any closer than 50 kilometers from the border between East and West Germany. And no more incidents of that kind occurred.}\footcite[p.~256]{khrushchev_memoirs_2007} His son Sergei, who annotated the memoir, admitted that this statement was false: \enquote{In practice such incidents occurred again from 1961 onward, but instead of the U-2 the Americans now used various types of Phantom or SR-71.}\footcite[p.~258]{khrushchev_memoirs_2007} On at least two occasions after the U-2 Incident, an American overflight crossed over into Soviet territory and the Soviets responded with... a diplomatic protest. The US quickly apologized, and that was that.\footcite{orlov_u-2_2007}

@@

% \subsection{Bridge of spies}
% The incidents involving U-2 aircraft during the Cold War perfectly mirror the traditional consequences for espionage because the United States succeeded in applying espionage norms to its use. Eisenhower's insistence on having the CIA manage Project Aquatone had the intended effect: the spy plane was, in story as well as fact, a civilian intelligence operation.

@@

% Both heads of state made clear in their writings that they considered Powers to be spy. On capturing Gary Powers, Khrushchev wrote: ``This was a hostile act by the leaders of the U.S. government, and they made no attempt to conceal it. They didn't think we had the capability of \textelp{} acquiring irrefutable proof that the United States was using methods that were impermissible in peacetime.''\footcite[p.~239]{khrushchev_memoirs_2007} Sergei's annotations once again contradicted his father's claim that these methods were impermissible; more than forty American reconnaissance aircraft were shot down by the USSR during the Cold War. As is often the case with espionage, Khrushchev wanted to make it seem like he was standing up for the integrity of the Soviet state without endangering the espionage equilibrium.

% Eisenhower, always an insightful thinker on issues of intelligence, understood that beneath the grandstanding, the fundamentals of international espionage still applied. In his own words:

% % \begin{quote}
% 	Furthermore, it seemed to me that Mr. Khrushchev's outbursts were hypnotizing the world with a passionate but highly distoreted presentation of one particular phase of international espionage. His government had been so notoriously involved in spying, especially in the United States, as to dwarf our activities, but by separating this particular type of espionage from all others he hoped to make convincing his charge of \enquote{warmongering.} To claim that because the equipment employed was an airplane with a camera, and therefore provocative of war, was plain silly, and I felt it necessary that the matter be put in perspective. \\

% 	The real issue at stake was not the fact that both sides conducted intelligence activities, but rather that the conduct and announced intentions of the Communists created the necessity for such clandestine maneuvers. As a consequence, the West---more specifically the Untied States as the major military power of the West---had to maintain constantly the capacity to detect any possible prelude to an infinitely more destructive Pearl Harbor.\footcite[p.~551]{eisenhower_waging_1965}
% % \end{quote}
% The next chapter examines in depth how the fear of a nuclear Pearl Harbor affected the entire American intelligence strategy, but even in the limited context of the spy plane, it is clear that Eisenhower deployed espionage to de-escalatory effect. Khrushchev had convinced the world that the Soviet missile capabilities were far more advanced than they actually were. Eisenhower was under tremendous pressure from the public, Republicans in Congress, and even his own staff to increase the defense budget. The only thing standing between Eisenhower and an accelerated arms race were detailed photo-reconnaissance reports from the U-2 that showed the Soviets were developmentally far behind what their public statements suggested.

% The Powers flight and Project Aquatone channel the norms of espionage in one final, but crucial respect: their strategic value easily outweighed the consequences of being caught. No one understood this better than President Eisenhower, who, when he was occasionally asked about the wisdom of the U-2 flights, would always reply: \enquote{Would you be ready to give back all of the information we secured from our U-2 flights over Russia if there had been no disaster to one of our planes in Russia?} In his telling, he never received an affirmative response.\footcite[p.~559]{eisenhower_waging_1965}

>>>>

\begin{table}[ht]
\centering
\begin{tabular}{llr}
\textbf{Date}     & \textbf{\makecell[l]{U.S. Service \&\\ Aircraft Type}}   & \textbf{General Location} \\
8 April 1950      & USN PB4Y2 Privateer           & Baltic Sea                           \\
6 November 1951   & USN P2V Neptune               & Sea of Japan                         \\
13 June 1952      & USAF RB-29                    & Sea of Japan                         \\
7 October 1952    & USAF RB-29                    & East of Hokkaido/Kuril Is.           \\
29 July 1953      & USAF RB-50                    & Sea of Japan                         \\
4 September 1954  & USN P2V Neptune               & Sea of Japan                         \\
7 November 1954   & USAD RB-29                    & East of Hokkaido.Kuril Is.           \\
18 April 1955     & USAF RB-47                    & Off Kamchatka Peninsula              \\
10 September 1956 & USAF RB-50                    & Sea of Japan                         \\
2 September 1958  & USADF C-130                   & Soviet Armenia
\\
1 May 1960        & CIA U-2                       & Sverdolsk, USSR                      \\
1 July 1960       & USAF RB-47                    & Barents Sea                          \\
10 March 1964     & USAF RB-66                    & East Germany
\end{tabular}
\caption{Summary of Soviet Shootdowns, 1950-1964}
\label{soviet-shootdowns}
\end{table}

@@

The general attitudes displayed in response to the Baltic Incident became the normal course of response to Soviet flights. The shootdown of the Navy Privateer was not an isolated incident; from then until the U-2 Incident in 1960, the United States lost nine more aircraft to Soviet air defenses. Each time, the pattern of responses was the same---each side released dueling statements of protest, and the situation continued more or less as before. A B-29 that went missing on October 7, 1952 hit the front page of the \emph{New York Times} two days later. This mission is on Peterson's list as a ferret flight. Farquhar notes that the incident happened during the US election season and that the media thought the attack was intended to lower American prestige, but he does not mention any diplomatic consequences.\footcite[p.~43-44]{farquhar_aerial_2015} The official US response, however, is preserved on the State Department website. As with the Baltic Incident, the US simply demanded that the Soviets pay for the plane and return any survivors.\footcite{the_new_york_times_u.s._1952} Two more shootdowns in March 1953---one involving an American F–84 Thunderjet and the other an RAF Lincoln bomber---actually resulted in conciliatory responses from both sides, including a secret meeting to reduce aerial tensions \enquote{of which little is known.}\footcite[p.~45]{farquhar_aerial_2015}

@@

I contend that the lack of espionage equipment is precisely the reason the United States made a sustained public effort to recover the four airmen, which is something that almost never happened with ferret flights.\footnote{While some of the flights were shot down overseas, making recovery difficult, I am referring here to situations where the US was aware that the Soviets likely held their airmen in captivity and still made little effort to get them back.} Farquhar summarizes the American response to this incident as follows: \enquote{Responding to the press attention, the Truman Administration acted swiftly, attempting to gain the fliers release through diplomatic pressure. The President ordered the Hungarian consulates in New York and Cleveland closed and banned private travel to the country. Legislatively, Truman asked Congress to pass a \$100 million Mutual Security Act to aid \enquote{selected persons residing in Soviet bloc states or refugees who wanted to form armed units} in opposition to Communism.}\footcite[p.~43]{farquhar_aerial_2015} Truman immediately took action against material Hungarian interests in the United States and threatened to empower \emph{armed dissidents} if the prisoners were not released. The contrast between this response and the one from just a year prior, when the US demanded \enquote{a prompt and thorough investigation} of the Baltic Incident, is stark.

@@

\subsection{Inventing the spy plane}
While Eisenhower was able to defuse some of the tension surrounding the military reconnaissance flights, even the most aggressive overflights could only hope to skirt the borders of Soviet territory; the United States couldn't just fly a B-50 over Moscow. What US intelligence-gathering capabilities needed was some way of photographing suspicious installations deep within Russia while still being able to claim the same benefit-of-the-doubt status that prevented even the most contentious overflights from escalating into a larger conflict. What they needed was a spy plane.

For the spy plane to serve its intended purpose as a minimally provocative agent of espionage then, two things must be true: its operation must be understood by adversaries to be non-military, and that distinction must meaningfully alter their threat perception and associated response. To build such a aircraft, Eisenhower approved Project Aquatone---what became the U-2. From its conception, the U-2 was purpose-built to be read as a civilian aircraft. Eisenhower required that the pilots of U-2 aircraft be CIA employees, specifically forbidding uniformed Air Force officers.\footcite[p.~33, Though many of the pilots did have Air Force backgrounds.]{lindgren_trust_2000} The CIA also had operational control of the program instead of the Air Force. \enquote{I want this whole thing to be a civilian operation,} Eisenhower wrote, when settling an operational dispute between the two departments. \enquote{If uniformed personnel of the armed services of the United States fly over Russia, it is an act of war---legally---and I don't want any part of it.}\footcite[p.~60. The original source for this quote is an \emph{OSA History} that requires codeword clearance. It is quoted here by the History Staff of the CIA.]{pedlow_central_1992} This allowed the United States to \enquote{truthfully deny,} in the phrasing of the CIA, that any US \enquote{military planes} had flown over the USSR---which they had to do when the inevitable Soviet protest notes were filed after U-2 overflights began in 1956.\footcite[p.~109]{pedlow_central_1992}

The steps Eisenhower took to ensure the U-2 civilian's character went beyond mere deception. If the only goal had been to achieve a level of plausible deniability in the event of a shootdown, then simply not wearing an Air Force uniform during the mission likely would have been sufficient. The \enquote{weather reconnaissance} excuse had even been used before with the ferret flights.\footcite[p.~45]{farquhar_aerial_2015} At the highest levels of American government, however, policymakers didn't just want to pretend that Aquatone was a civilian operation, they wanted it to \emph{be} a civilian operation.\enquote{It is of utmost importance to differentiate in our minds, and to cause the Russians to differentiate in theirs, between Aquatone-type operations and reconnaissance by military aircraft} read a top-secret CIA memo from 1956.\footcite[p.~1]{miller_suggestions_1956} Both halves of that equation were equally important. The Russians must believe that these operations are peacetime civilian operations, and they must be correct about it.

At first, what made the U-2 flights \enquote{count} as espionage appears to be a deeply unsatisfying tautology. They counted as espionage because they were espionage! But Eisenhower's precautions were based in the reality of the United State's previous experience with reconnaissance flights. He had already been testing the limits of what the Soviet Union would be willing to accept. No finite list of attributes could ensure that the Russians wouldn't treat the U-2 plane---a massive leap in military surveillance technology---as a military development. But if the people in charge of Project Aquatone could convince \emph{themselves} that it was a civilian project, then they stood the best chance of having the Russian interpret it that way. The U-2 simply had to be a spy plane, in a manner that is both instantly recognizable and ontologically undefinable.

These precautions had the intended effect, even though at first it seemed like the Soviets were going to extract significant concessions. A Soviet Colonel in the USSR Air Defense later recalled that Khrushchev ``clearly viewed the violation of their nation's skies by a foreign reconnaissance aircraft \textelp{} as a political provocation'' and Khrushchev played it up.\footcite{orlov_u-2_2007} Three days after NASA quietly reported that they had lost a U-2 type weather reconnaissance plane, Khrushchev announced that the USSR had shot down an American plane that had flown into their airspace in \enquote{an aggressive provocation.}\footcite[p.~112]{powers_operation_2004} When the Americans stuck to their cover story, Khrushchev gleefully debunked it by revealing that they had taken Powers alive, Eisenhower admitted that the Gary Powers had been flying a spy plane, in a press conference where Eisenhower memorably declared that espionage was a \enquote{a distasteful but vital necessity} to prevent another attack like Pearl Harbor.\footcite{eisenhower_news_1960}

An yet few real consequences followed. No one disputes that the collapse of the Paris Summit was directly attributable to the Gary Powers shootdown but, frankly, summits just aren't that important. From 1953 to the end of the Cold War in 1991, there were 23 total US-USSR summits.\footcite{fain_chronology_2011} The main issue that did not get resolved in 1960 was the status of Berlin. It also did not get resolved at the next summit in 1961, a comparatively warmer affair between Khrushchev and Eisenhower's successor, John F. Kennedy. If the worst that happened because of the U-2 incident was an unproductive summit meeting, then that is a relatively minor punishment. With the late rollout of the captured pilot, the dramatic scene at the summit, and the cancellation of Eisenhower's visit, Khrushchev publicly embarrassed the United States without meaningfully altering Soviet policy towards it.

\subsection{Evaluating the consequences of aerial espionage}
The U-2 Incident is often cited as a seminal failure of the Eisenhower administration. In the traditional telling, it significantly worsened US-Soviet relations at time when it seemed like lasting peace might be possible. I disagree. Instead, the U-2 Incident should be understood as an embarrassing but otherwise typical failure of an intelligence operation, one consistent with the minimal consequences typically applied to espionage.\footnote{While I argue that the response to the U-2 flight was within the reasonable bounds for espionage, I acknowledge that it was still unprecedented in the context of reconnaissance flights. There are a number of reasonable explanations for why the U-2 shootdown escalated. Partially the Soviets felt humiliated by a technology for which they had no countermeasures and the uniquely insulting way the overflights flew deep into Soviet territory. Eisenhower thought that Khrushchev might have been looking for an excuse to cancel his visit to Russia. See: \cite[p.~555]{eisenhower_waging_1965}. Others in his administration believed that the U-2 Incident was an excuse to justify an intensification of hostilities that had already been decided by hard-liners in the Kremlin. See: \cite[p.~328]{kistiakowsky_scientist_1976}.} What would a major punishment have looked like? One proportional response Khrushchev could have considered was to force the issue on regional air bases. Since the US was using nearby bases to launch invasive reconnaissance missions, the USSR could have demanded that the US either close some of them or agree to halt surveillance flights. This likely would have involved some sort of trade, but it is not an unreasonable hypothetical. The Soviets were incredibly preoccupied with removing the forward-deployed Jupiter missiles from Turkey. Although the missiles were deployed to bases well within NATO territory, and arguably less threatening to Soviet security than American overflights of the Soviet bloc, the existence of these missile became a central bargaining chip in the Cuban Missile Crisis. By contrast, the Soviets simply did not have the same level of interest in curtailing reconnaissance flights. That both sides of the Cold War had strong incentives to avoid military escalation is well-documented, but at least threatening military retaliation for launching provocative military operations would not seem to have been out of place.

Most scholars who study this period argue that the cumulative tension created by these flights contributed to an overall chilly tone in US-Soviet relations. Farquhar, for instance, refers to this as the \enquote{cycle of hostility,} in which the series of aerial incidents increased suspicions of military buildup on both sides, leading to even more aerial incidents as both powers attempted to verify their suspicions.\footcite[p.~43]{farquhar_aerial_2015} While it's true that the there were many aerial incidents, it is not at all clear how this affected any other aspect of US-Soviet diplomacy. I can't definitively say that there wasn't some diplomatic issue that would have gotten resolved at a summit absent these reconnaissance flights, but I can't locate an obvious one either.

Consider the period of time over which these flights took place. If aerial reconnaissance had risen to anything more than a lingering irritant, then it should have to have been resolved in some way before 1960. The US-Soviet relationship saw huge shifts, including periods of optimism, over the 11 years preceding the U-2 incident. During this period, hundreds of flights took place in contentious territory---many of them purposefully violating Soviet boundaries, causing some of them to be shot down---and the most significant consequence was a failed summit meeting, after which the flights continued on almost as before. Both sides were so willing to contain the issue that many of the airmen never saw their families again.

Until the day he died, Khrushchev greatly exaggerated the lasting effect of the U-2 Incident. He wrote in his memoir that \enquote{the commander of American forces in West Germany gave the order not to fly any closer than 50 kilometers from the border between East and West Germany. And no more incidents of that kind occurred.}\footcite[p.~256]{khrushchev_memoirs_2007} His son Sergei, who annotated the memoir, admitted that this statement was false: \enquote{In practice such incidents occurred again from 1961 onward, but instead of the U-2 the Americans now used various types of Phantom or SR-71.}\footcite[p.~258]{khrushchev_memoirs_2007} On at least two occasions after the U-2 Incident, an American overflight crossed over into Soviet territory and the Soviets responded with... a diplomatic protest. The US quickly apologized, and that was that.\footcite{orlov_u-2_2007}

% As is characteristic of espionage, when the US really needed intelligence that only an overflight could provide, the diplomatic consequences that the USSR had imposed were not sufficient to discourage it. Two months to the day after Powers was shot down, a Soviet MiG-19 opened fire on an American RB-47H and captured the plane's navigator and co-pilot. While Powers was serving a 10-year sentence in Soviet prison, the two airmen were returned to the US in less than six months with only mild fanfare. Khrushchev wrote in his memoir that he wanted to continue \enquote{our general line of peaceful coexistence} and cited the resolution of this incident as an example of such, because the US was forced to make a formal request for the return of the airmen.\footcite[p.~256-257]{khrushchev_memoirs_2007} The only concessions the US made in return were to announce the discontinuation of its U-2 overflights (which Kennedy was already committed to) and to not make an issue of the illegal detention of the pilots.\footcite{time_cold_1961} For spy flights after Powers, the USSR simply took the necessary immediate countermeasures and shot down the planes that it could. Just as with traditional espionage, the countermeasures did not punish the attempt.

@@

% The incidents involving U-2 aircraft during the Cold War perfectly mirror the traditional consequences for espionage because the United States succeeded in applying espionage norms to its use. Eisenhower's insistence on having the CIA manage Project Aquatone had the intended effect: the spy plane was, in story as well as fact, a civilian intelligence operation.

@@

Both heads of state made clear in their writings that they considered Powers to be spy. On capturing Gary Powers, Khrushchev wrote: ``This was a hostile act by the leaders of the U.S. government, and they made no attempt to conceal it. They didn't think we had the capability of \textelp{} acquiring irrefutable proof that the United States was using methods that were impermissible in peacetime.''\footcite[p.~239]{khrushchev_memoirs_2007} Sergei's annotations once again contradicted his father's claim that these methods were impermissible; more than forty American reconnaissance aircraft were shot down by the USSR during the Cold War. Khrushchev wanted to make it seem like he was standing up for the integrity of the Soviet state without endangering the espionage equilibrium.

Eisenhower, an insightful thinker on issues of intelligence, understood that beneath the grandstanding, the fundamentals of international espionage still applied. \enquote{[Khrushchev's] government,} he wrote, \enquote{had been so notoriously involved in spying, especially in the United States, as to dwarf our activities, but by separating this particular type of espionage from all others he hoped to make convincing his charge of \enquote{warmongering.} To claim that because the equipment employed was an airplane with a camera, and therefore provocative of war, was plain silly, and I felt it necessary that the matter be put in perspective.} \footcite[p.~551]{eisenhower_waging_1965}

The Powers flight and Project Aquatone channel the norms of espionage in one final, but crucial respect: their strategic value easily outweighed the consequences of being caught. No one understood this better than President Eisenhower, who, when he was occasionally questioned about the wisdom of the U-2 flights, would always reply: \enquote{Would you be ready to give back all of the information we secured from our U-2 flights over Russia if there had been no disaster to one of our planes in Russia?} In his telling, he never received an affirmative response.\footcite[p.~559]{eisenhower_waging_1965}
